---
title: "fitting_LPA_model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fitting_LPA_model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
load the dpm.lpa R package
```{r setup}
library(dpm.lpa)
# To build: devtools::build_vignettes()
```
update relevant Python packages to the latest versions (this is optional and thus commented out)
```{r}
# update_python_packages()
```
### import and standardize data
```{r}
data(iris)#read.csv("data/LPA example data.csv", row.names = 1)
x = data.frame(scale(iris[, c('Sepal.Width', 'Petal.Width')]))
```
### fit the DPM-LPA model
```{r, warning=FALSE}
fitted_lpa = fit_dpm_lpa(x = x,
                         restarts = 10)
```
## look at model output

### print summary of profile fit
```{r print_summary, warning=FALSE}
print_lpa_summary(fitted_lpa)
```
### plot residuals
```{r residuals, warning=FALSE, fig.show='hide'}
plot_lpa_residuals(fitted_lpa)$show()
```

### plot estimated profile means (mu)
```{r profile_means, warning=FALSE}
plot_profile_means(fitted_lpa)$show()
```
### plot estimated profile base rates (probabilities, pi)
```{r profile_base_rates, warning=FALSE}
# FIX THIS (START FROM 1 NOT WORKING; need to check that the Python code works properly, then upload that to PyPi, then update the downloaded version here)
plot_profile_probs(fitted_lpa, start_profile_labels_from1 = FALSE)$show()
```
### analyze similarity between profiles
```{r profile_similarity, warning=FALSE}
profile_similarity(model = fitted_lpa)$plot$show()
```
## outcome analysis

We now want to see if the latent profiles are related to some outcome variables.
We first use the 'outcome_analysis_normal' function to run the analysis. This
returns a list with various results. We save this list as 'result'. You can view the
function's help file (using '?outcome_analysis_normal') for a description of
each of the elements contained in the function's output. We are mainly interested
in two outputs: the results of analysis using Bayes factors, and a plot of estimated
profile outcome means.
```{r run_outcome_analysis, warning=FALSE}
result = outcome_analysis_normal(model = fitted_lpa,
                                 #y = data.frame(matrix(rnorm(n = 2*150), nrow = 150, ncol = 2)))
                                 y = data.frame(y1 = rnorm(n = 150), y2 = rnorm(n = 150)))
                        #y = data.frame(scale(as.numeric(iris$Sepal.Length))))
```
### Bayes Factors
```{r bayes_factors, warning=FALSE}
result$bayes_factors
```
### Plot of profile outcome variable means (95% credible interval)
```{r outcome_plot}
result$plot$show()
```

```{r}
hist(rnorm(100))
```

